Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-05-04 src/tex/ms.tex
[showyourwork] overleaf sync

% Filler text
\usepackage{blindtext}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage{amssymb}
\let\tablenum\relaxgit 
\usepackage{siunitx}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\arcsinh}{arcsinh}
\newcommand{\chandra}{\textit{Chandra}~}
\newcommand{\xmm}{\textit{XMM}~}
\newcommand{\fermi}{\textit{Fermi}-LAT~}
\newcommand{\jolideco}{\textit{Jolideco}~}
\newcommand{\aposteriori}{a~posteriori~}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\xray}{X-ray\xspace}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}\PackageWarning{TODO:}{#1!}}

@@

% Abstract with filler text
    \begin{abstract}
        We present a new method for Joint Likelihood Deconvolution (Jolideco) of a set of astronomical observations of the same sky region in the presence of Poisson noise. The method reconstructs a single flux image from a set of observations by optimizing the \aposteriori joint Poisson likelihood of all observations under a patch based image prior. The patch prior is parameterised by a Gaussian Mixture model (GMM) and learned from astronomical images with high signal to noise ratio, such as the James Webb Telescope. During the reconstruction process the patch prior adapts  to the structures in the data by finding the most likely GMM component for each patch in the image. By applying the method to simulated data we show that both the combination of multiple observations as well as the patch based prior lead to an much improved reconstruction quality in many different source scenarios as well as signal to noise regimes. We show also that the method yields superior reconstruction quality to alternative standard methods such as the Richardson-Lucy method. We also show that the method can be highly adapted to various analysis scenarios by learning problem specific GMM prior  such as images of extragalactic or Galactic sources.
    \end{abstract}

@@

    \section{Introduction}
    The resolution of any physical or astronomical imaging process is inherently limited by the instrument or telescope used. Furthermore, the quality of the resulting image is affected by background or instrumental measurement noise and non-uniform exposure. This is especially true in the low signal-to-noise regime, where images are affected by Poisson noise. This includes images taken by \xray and \gammaray~telescopes but also images obtained with positron emission tomography or single-photon emission computed tomography.

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is called de-blurring or de-convolution, as the the data is obtained via a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is called inverse problems. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as un-blind de-convolution. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant activate galactic nuclei or a model of the PSF can obtained from simulation of the telescope or instrument.

    \subsection{Related Work}
    In literature there have been many efforts to solve the un-blind deconvolution problem for images affected by Poisson noise. Most methods rely fundamentally on modeling the image generating process as a linear process. Given a known point spread function a prediction of the discrete measured image can be obtained via \textit{forward folding} and be compared against an observation assuming Poisson statistics per pixel. However the methods differ in their additional prior or regularisation assumptions as well as their approach to optimization or sampling of the underlying likelihood function.

    The standard baseline method for image de-convolution was proposed independently by \cite{Richardson1972} and \cite{Lucy1974} (we will further refer to this method as \textit{RL}). The method involves estimating the original image by iteratively convolving the observed image with the PSF and dividing by the result of convolving the estimated image with the transposed PSF. The iterations continue until the estimated image converges to a satisfactory solution.

@@

    They showed that the patch based GMM prior led to much improved image reconstructions for multiple inverse problems such as de-noising, de-blurring and inpainting.

@@

    \subsection{Some Remarks on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the \textit{de facto} standard for de-blurring and super-resolution of natural images \todo{reference review?}. However their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical x-ray and \gammaray data.

@@

    Some of the challenges of de-blurring with CNN are due to the convolutional nature of the networks. In a single layer, linear approximation, the network would need to learn the inverse filter of the PSF, corresponding to the well known \textit{Wiener filter} solution of the problem. This requires large convolution kernels or deep network architectures. \cite{Li2014} have introduced a baseline architecture for a deep CNN for image de-blurring by introducing large kernels under the separability assumption.

@@

    Or even an arbitrary combination of the possibilities listed above.

@@

    \subsection{Patch Based Image Prior}
    In general it is difficult to capture global image statistics in a prior likelihood taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the number of pixels squared and thus quickly becomes computationally untraceable. However astronomical images typically only show local correlations between pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light (structure in the CMB might be an exception here). This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the same patch based image prior approach as introduced by \cite{Zoran2011}. We use a pre-trained Gaussian Mixture Model (GMM) to evaluate and approximate a patch based prior for the reconstructed image $\mathbf{x}$.

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch
    from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the GMM component becomes a discrete hyper-parameter in the model that is optimized along with the image $\mathbf{x}$.

@@

    \subsubsection{Image Normalisation and Dynamic Range}
    Astronomical images typically show a much higher dynamic range compared to 
    natural images. This is mostly due to the existence of point sources, which
    rarely occur in natural images. Astronomical point sources typically relate
    to compact bright objects, such as active galactic nuclei at high distances,
    or binary objects and stars at Galactic distances.

@@

    \begin{figure}
        \script{image-norms.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/image-norms.pdf}
            \caption{
                Various image norms with different dynamic behaviour.
            }
            \label{fig:image-norms}
        \end{centering}
    \end{figure}

@@

    \subsection{Cross Validation}
    Cross validation (CV) is an established method for model selection in statistics
    and machine learning to prevent over-fitting the data. To perform CV the available
    data is slit into a "training" and "validation" dataset. While optimizing the
    log-likelihood value is monitored

    The use of multiple dataset allows to use cross-validation to prevent overfitting.
    One of the standard techniques in "machine learning"...

@@

@@

    \subsubsection{Statistical Errors}
    \begin{itemize}
        \item Bootstrapping observations
        \item Bootstrapping events
        \item Hessian diagonal
    \end{itemize}

@@

    \subsubsection{Zoran \& Weiss}
    As baseline reference GMM we use the original model provided by \cite{Zoran2011}. The model was learned from the Berkley image database \todo{reference}, which contains natural images and image and every day scenes and objects. The model was learned by splitting the images into 8x8 patches and provided by the authors as part of the source code for download\footnote{\url{https://people.csail.mit.edu/danielzoran/epllcode.zip}}.
    More details of how the model was learned are given in the original publication.

@@

    \subsubsection{JWST Cas A Image}
    As a last GM prior we use the promotional image of JWST realeased by...
    The image is particularly suited becasue of the excellent angular resolution, extrating many patches for the same object
    as well as the high signal to noise ratio.

    \subsubsection{NRAO Jets, Jets, Jets}
    \todo{Include the jet prior or not?}
    For the specific application of jet detection and analysis in extract galactic data we learned a GMM patch prior from radio jet images. It represents the prior knowledge of both the morphology and preferred direction of the jet. As reference data we chose
    a PR image released under the name \textit{Jets, jets jets}  by the NRAO collaboration containing ~150 example jets from their data\footnote{\url{https://public.nrao.edu/gallery/quicklook-samples/}}.

    We first split the compiled image into 150 individual examples. For each image we ran a blob detection algorithm and line fit to find the preferred direction of the jet. Using this information we aligned all images to the same horizontal reference direction of the jet split all images into patches of size 8x8 pixels. In total we extracted \todo{insert number} patches and used those to train a GMM with 32 components. The prior is useful in a scenario were the direction of the jet known from e.g. previous radio observations.

    Alternatively we used a \textit{data augmentation} approach to learn a GMM prior with a randomized direction of the jet. For this we rotated the jet image after alignment by 
    a random angle. Then again we split each image into patches of size 8x8 pixels. In total we created \todo{insert number} patches and learned a GMM with 64 components. The prior
    is adapted to the analysis scenario of jet detection, where no prior knowledge on the
    direction is available.

@@

    \begin{table*}[t]
      \centering
      \begin{tabular}{l|cr}
        \hline
        Scenario & Description & Parameters \\
        \hline
        A1 & A grid of point sources & x, y \\
        B1-3 & Extended Gaussian shaped source surrounded by point sources & $\sigma$ \\
        C1-3 & Point Sources on top of a disk shaped source, jet like line features & $r_{\mathrm{Disk}}$ \\
        D1-5 & A spiral-arm structure surrounded by point sources and a large ring like feature & 9 \\

@@

    \todo{Vinay writes this?}
    \vlk{yes}

    \vlk{We have devised four distinct arrangements of combinations of point and extended sources of different shapes to test the algorithms.  The arrangements are 
    \begin{itemize}
        \item[(A)] {\bf asterism:} A set of four point sources arranged around an extended Gaussian source
        \item[(B)] {\bf points:} Point sources arranged with different separations
        \item[(C)] {\bf shield:} Disk shaped flat extended source with superposed point sources and linear jet-like features
        \item[(D)] {\bf spiral:} Extended thin double-spiral with a flat disk at the center and point sources adjacent to it
    \end{itemize}
    The arrangements are shown in Figure~TBD, and the details of the locations and brightness of each component is listed in Table~TBD.}

    \begin{figure*}
        \centering\includegraphics{figures/scenarios.pdf}
        \caption{Illustrating the source patterns used to make test datasets.  The patterns shown are for the {\sl (A)} {\tt asterism}, {\sl (B)} {\tt points}, {\sl (C)} {\tt shield}, and {\sl (D)} {\tt spiral} cases (see Section~\ref{sec:testmodels}), convolved with a sharp PSF (Gaussian with $\sigma=2$~pix).  All images are of size $128{\times}128$.  The dashed red and magenta boxes depict $32{\times}32$ and $64{\times}64$ regions of interest.  Note that the boxes are displaced to the lower left corner in Case B, and leftwards in Case D.
        convolved with a diffuse PSF (Gaussian with $\sigma=6$~pix), \todo{add back boxes if we have time}.
        }
        \label{fig:simulated_models_chandra}
    \end{figure*}

    We use the test datasets provided by Vinay et al.
    First we evaluate the performance of the method on a set of simulated observations.
    For this we assume:

    \begin{itemize}
        \item An instrument with good angular resolution, but low effective area (e.g. like Chandra)
        \item An instrument with worse angular resolution, but higher effective area (e.g. like XMM)
    \end{itemize}

    For both scenarios we assume a Gaussian PSF of sizes $\sigma = 2$ pixels and  $\sigma = 5$.
    As true flux we consider the following scenarios:

    \begin{itemize}
        \item Point source with varying distances and brightness
        \item Disk and point sources of varying brightness
        \item Spiral and point sources of varying brightness
    \end{itemize}

    Background levels of $\lambda_{Bkg}= 0.001, 0.01 \textrm{and} 0.1 \textrm{counts/pixel}$.

    Show results of experiments on simulated toy datasets

@@

        \item \textbf{Jolideco (Patch, GLEAM v0.1)}: \jolideco with a patch prior learned on GLEAM data, described in Section~\ref{sssec:gleam-radio-data}. It also uses the standardised patch normalisation as described by Equation~\ref{eq:patch-norm}. We use cycle spinning, a stride of 4 pixels and 1000 iterations. Which optimizes "to the end" until there is no relevant change in the posterior log-likelihood. In addition we use and upsampling factor of 2.

@@

    We ran the comparison on the whole parameter grid of methods, source scenarios, background levels and instrument scenarios. However in the following we only discuss a subset of the results, which we consider the most relevant. All results are available at \url{https://jolideco.github.io/jolideco-comparison}.

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \script{comparison-scenarios.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of different deconvolution methods for different selected scenarios, as described in Section~\ref{subsec:test-datasets}. The simulation of the data used a fixed background level of, a constant exposure and a Gaussian PSF of \todo{x pixels}, corresponding to the "Chandra" scenario. The left column shows the simulated counts data and the second most column from the left the underlying ground truth to compare. The other column show the reconstructed flux by the different methods and prior assumptions. The methods are described in detail in Section~\ref{subsec:methods}.
                A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of methods on the different source scenarios.
    Figure \ref{fig:comparison-scenarios} shows the comparison of the methods for a selected subset of scenarios A1, B3, C3 and D4.

@@

    For scenario B3 all method recover the point sources and the central extend source. 
    However large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected the uniform prior decompose the emission into many point sources, neglecting the spatial correlation between them. \textbf{Pylira} shows slightly improved results with the gaps between the bright emission peeks being more filled, however still lacks a convincing result into reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. The GLEAM v0.1 patch prior reconstructs the 
    sources almost perfectly, visually indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a visible extension in the surrounding point sources.

    Scenario C3 challenges the methods to reconstruct a weak extended source under four brighter point sources. While again all methods reconstruct the point sources well, the visual differences are large for the weak underlying extended emission. The uniform prior with 1000 iterations

    \jolideco shows excellent performance of a variety of morphological shapes. 
    it converges to the reference solution for very high signal to noise.

    \par{\textbf{Pylira}}
    Pylira shows 
    \begin{itemize}
        \item Shows good results for point sources
        \item Slightly improved results on extended source, but still "speckly"
        \item Pixels loose spatial correlation over larger scales..
    \end{itemize}

    \par{\textbf{Jolideco, Uniform Prior n=1000:}}
    The uniform priors show the expected behaviour: decompose into point sources, when "optimized to the end".

    \begin{itemize}
        \item the uniform priors show the expected behaviour: decompose into point sources, when "optimized to the end".
        \item yields very good results, if the true underlying source are only point sources
        \item results are very bad, once extended emission is present. Result is "speckly" with no visible spatial correlation between nearby pixels.
    \end{itemize}

    \par{\textbf{Jolideco, Patch Prior GLEAM}}
    \begin{itemize}
        \item shows best results
        \item slightly worse results on point sources only
        \item visually very close to ground truth (!)
        \item good compromise between sufficient separation of point sources and good reconstruction of extended sources.
    \end{itemize}

    \subsubsection{Comparing Signal-to-Noise Scenarios}
    %
    \begin{figure*}
        \script{comparison-signal-noise.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-signal-noise.pdf}
            \caption{
                Comparison of the \jolideco method with the GLEAM v0.1 prior for different background levels and signal strength represented by scenario D1-5. The images are grouped by background levels with each column group showing the counts data in the left column and the reconstruction in the right column. The rows show the scenarios D1-5 as described in Section~\ref{subsec:test-datasets}. The background levels correspond to \qty[mode = text]{0.001}{cts / pixel}, \qty[mode = text]{0.01}{cts / pixel} and \qty[mode = text]{0.1}{cts / pixel}. The complete results for the simulations and additional information are available at \url{https://jolideco.github.io/jolideco-comparison}. \todo{the background handling does not yet seem right...maybe we should let the background norm free}
            }
            \label{fig:comparison-signal-noise}
        \end{centering}
    \end{figure*}
    %
    As a second experiment we explored the dependency of the results on background level and signal strength. Figure~\ref{fig:comparison-signal-noise} shows the results for the \jolideco method with the GLEAM prior in the scenario D1-5. As expected the reconstruction quality improves with signal strength. The method seems to converge to the ground truth for very high signal-to-noise ratio.

    Es expected once the signal strength becomes very large there is no dependency of the result on the background level anymore (scenario D5)

    \todo{We should give a single number for the S/N, how to compute? Per source, like spiral ring and point sources...}

    Some observations:

>>>>

% Filler text
\usepackage{blindtext}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage[autostyle]{csquotes}
\usepackage{amssymb}
\let\tablenum\relaxgit 
\usepackage{siunitx}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\arcsinh}{arcsinh}
\newcommand{\chandra}{\textit{Chandra}~}
\newcommand{\xmm}{\textit{XMM}~}
\newcommand{\fermi}{\textit{Fermi}-LAT~}
\newcommand{\jolideco}{\textit{Jolideco}~}
\newcommand{\aposteriori}{a~posteriori~}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\xray}{X-ray\xspace}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}\PackageWarning{TODO:}{#1!}}

@@

% Abstract with filler text
    \begin{abstract}
        We present a new method for Joint Likelihood Deconvolution (Jolideco) of a set of astronomical observations of the same sky region in the presence of Poisson noise. The method reconstructs a single flux image from a set of observations by optimizing the \aposteriori joint Poisson likelihood of all observations under a patch based image prior. The patch prior is parameterised by a Gaussian Mixture model (GMM) and learned from astronomical images with high signal to noise ratio, such as the James Webb Telescope. During the reconstruction process the patch prior adapts  to the structures in the data by finding the most likely GMM component for each patch in the image. By applying the method to simulated data we show that both the combination of multiple observations as well as the patch based prior lead to a much improved reconstruction quality in many different source scenarios as well as signal to noise regimes. We show also that the method yields superior reconstruction quality to alternative standard methods such as the Richardson-Lucy method. We also show that the method can be highly adapted to various real analysis scenarios by learning problem specific GMM prior  such as images of extragalactic or Galactic sources.
    \end{abstract}

@@

    \section{Introduction}
    The resolution of any physical or astronomical imaging process is inherently limited by the instrument or telescope used. Furthermore, the quality of the resulting image is affected by background or instrumental measurement noise and non-uniform exposure. This is especially true in the low signal-to-noise regime, where images are affected by Poisson noise. This includes images taken by \xray and \gammaray~telescopes but also medical images obtained with methods such as positron emission tomography or single-photon emission computed tomography.

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is often called deblurring or deconvolution, as it aims to invert the effect of the data being degraded by a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is categorised as \textit{inverse problems}. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as unblind deconvolution. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant activate galactic nuclei or a model of the PSF can obtained from simulation of the telescope or instrument.

    \newpage
    \subsection{Related Work}
    In literature there have been many efforts to solve the unblind deconvolution problem for images affected by Poisson noise. Most methods rely fundamentally on modeling the image generating process as a linear process. Given a known point spread function a prediction of the discrete measured image can be obtained via \textit{forward folding} and be compared against an observation assuming Poisson statistics per pixel. However the methods differ in their additional prior or regularisation assumptions as well as their approach to optimization or sampling of the underlying likelihood function.

    The standard baseline method for image deconvolution was proposed independently by \cite{Richardson1972} and \cite{Lucy1974} (we will further refer to this method as \textit{RL}). The method involves estimating the original image by iteratively convolving the observed image with the PSF and dividing by the result of convolving the estimated image with the transposed PSF. The iterations continue until the estimated image converges to a satisfactory solution.

@@

    They showed that the patch based GMM prior led to much improved image reconstructions for multiple inverse problems such as denoising, deblurring and inpainting.

@@

    \subsection{Some Remarks on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the \textit{de facto} standard for deblurring and super-resolution of natural images \todo{reference review?}. However their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical x-ray and \gammaray data.

@@

    Some of the challenges of deblurring with CNN are due to the convolutional nature of the networks. In a single layer, linear approximation, the network would need to learn the inverse filter of the PSF, corresponding to the well known \textit{Wiener filter} solution of the problem. This requires large convolution kernels or deep network architectures. \cite{Li2014} have introduced a baseline architecture for a deep CNN for image deblurring by introducing large kernels under the separability assumption.

@@

    Or even an arbitrary combination of the possibilities listed above. It could also be the combination of observations in different energy ranges, however in this case the assumption of the morphology not changing with energy would apply, which is rarely fulfilled for the astronomical data we are dealing with.

@@

    \subsection{Patch Based Image Prior}
    In general it is difficult to capture global image statistics in a prior likelihood taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the number of pixels squared and thus quickly becomes computationally untraceable. However astronomical images typically only show local correlations between pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light (structure in the CMB might be an exception here). This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the idea of a patch based image prior approach as introduced by \cite{Zoran2011}. We use a pre-trained Gaussian Mixture Model (GMM) to evaluate and approximate a patch based prior for the reconstructed image $\mathbf{x}$.

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch
    from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the most likely GMM component per patch, becomes a discrete hyper-parameter in the model that is optimized along with the image $\mathbf{x}$.

@@

    \subsubsection{Patch Normalisation and Dynamic Range}
    Astronomical images typically show a much higher dynamic range compared to 
    natural images. This is mostly due to the existence of point sources, which
    rarely occur in natural images. Astronomical point sources typically relate
    to compact bright objects, such as active galactic nuclei at high distances,
    or binary objects and stars at Galactic distances. In addition the visible matter in the universe is sparse such that astronomical images typically show sparse bright emission on a dark background.

    \cite{Zoran2011} used a patch normalisation where the mean of the patch is subtracted. This means the same morphological features are present in the GMM at various levels of contrast. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, only at the cost of introducing a lot more components to the GMM, which is unfeasible in practice due to limited availability of training data and computational resources.

@@

    % \begin{figure}
    %     \script{image-norms.py}
    %     \begin{centering}
    %         \includegraphics[width=\linewidth]{figures/image-norms.pdf}
    %         \caption{
    %             Various image norms with different dynamic behaviour.
    %         }
    %         \label{fig:image-norms}
    %     \end{centering}
    % \end{figure}

@@

    \subsection{Choice and Tuning of Hyperparameters}
    \subsubsection{Cross Validation}
    Cross validation (CV) is an established method for model selection in statisticd and machine learning to prevent a model from overfitting to the training data. To perform CV the available data is slit into a "training" and "validation" dataset. While optimizing the log-likelihood value is monitored  The use of multiple dataset allows to use cross-validation to prevent overfitting. One of the standard techniques in "machine learning"...

@@

    \subsubsection{Other Approaches}

    Self supervised Poisson loss? J-invariance has been shown to work, Grid Search? Calibration on similar data with ground truth?

@@

    \subsection{Statistical Errors}
    \begin{itemize}
        \item Bootstrapping observations
        \item Bootstrapping events
        \item Hessian diagonal
    \end{itemize}

@@

    \subsubsection{Zoran \& Weiss}
    As baseline reference we use the original GMM provided by \cite{Zoran2011}. The model was learned from the Berkeley image database \citep{Martin2001} which contains natural images and images of every day scenes and objects. The model was learned by splitting the training images into patches of size 8x8 pixels. This resulted in a training set of $2 \cdot 10^{6}$ patches, which were used to fit a GMM with 200 components. The model is provided by the authors as part of the source code for download\footnote{\url{https://people.csail.mit.edu/danielzoran/epllcode.zip}}. More details of how the model was learned are given in the original publication.

@@

    \subsubsection{JWST Cassiopeia A}
    The James Webb Space Telescope is the it delivers data a very high resolution and excellent signal to noise ratio. We extracted 2 million patches out of the image...
    Would be 
    Of \textit{Cassiopeia A}
    As a last GM prior we use the promotional image of JWST realeased by...
    The image is particularly suited becasue of the excellent angular resolution, extrating many patches for the same object
    as well as the high signal to noise ratio.

    Taken from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}

    % \subsubsection{NRAO Jets, Jets, Jets}
    % \todo{Include the jet prior or not?}
    % For the specific application of jet detection and analysis in extract galactic data we learned a GMM patch prior from radio jet images. It represents the prior knowledge of both the morphology and preferred direction of the jet. As reference data we chose
    % a PR image released under the name \textit{Jets, jets jets}  by the NRAO collaboration containing ~150 example jets from their data\footnote{\url{https://public.nrao.edu/gallery/quicklook-samples/}}.

    % We first split the compiled image into 150 individual examples. For each image we ran a blob detection algorithm and line fit to find the preferred direction of the jet. Using this information we aligned all images to the same horizontal reference direction of the jet split all images into patches of size 8x8 pixels. In total we extracted \todo{insert number} patches and used those to train a GMM with 32 components. The prior is useful in a scenario were the direction of the jet known from e.g. previous radio observations.

    % Alternatively we used a \textit{data augmentation} approach to learn a GMM prior with a randomized direction of the jet. For this we rotated the jet image after alignment by 
    % a random angle. Then again we split each image into patches of size 8x8 pixels. In total we created \todo{insert number} patches and learned a GMM with 64 components. The prior
    % is adapted to the analysis scenario of jet detection, where no prior knowledge on the
    % direction is available.

@@

    To evaluate the reconstruction performance of \jolideco and compare it against other methods we defined a set of test source scenarios. They consist of simplified elementary morphological structures resembling those found in astronomical images. We have devised four distinct arrangements of combinations of point and extended sources of different shapes. The scenarios are:

    \begin{itemize}
        \item[(A)] {\enquote{\bf Points}:} point sources arranged in a grid with different separations and brightness. 
        \item[(B)] {\enquote{\bf Asterism}:} a set of four point sources arranged around an extended Gaussian source
        \item[(C)] {\enquote{\bf Shield}:} disk shaped flat extended source with superposed point sources and linear jet-like features 4 different directions.
        \item[(D)] \enquote{{\bf Spiral}:} extended thin double-spiral with a flat disk at the center and point sources adjacent to it. 
    \end{itemize}

    In addition we consider two different instrument scenarios:

    \begin{itemize}
        \item {\enquote{\bf Chandra}:} an instrument with good angular resolution, but low effective area. We assume a Gaussian PSF of sizes $\sigma = 2$ pixels and a reference effective area of \qty[mode=text]{1}{A.U.}.
        \item {\enquote{\bf XMM}:} an instrument with worse angular resolution, but higher effective area. We assume a Gaussian PSF of sizes $\sigma = 6$ pixels and an effective area of \qty[mode=text]{5}{A.U.}.
    \end{itemize}

    The ground truth and corresponding expected counts for the combination of scenarios are shown in Figure~\ref{fig:scenarios}, and the details of the locations and brightness of each component is listed in Table~\ref{tab:scenarios}.

    \begin{figure*}
        \label{fig:scenarios}
        \centering\includegraphics{figures/scenarios.pdf}
        \caption{Illustration of the expected counts for the different source and instrument scenarios we used to make test datasets. The source patterns shown are for the {\sl (A)}  {\tt points}, {\sl (B) {\tt asterism}}, {\sl (C)} {\tt shield}, and {\sl (D)} {\tt spiral} cases (see Section~\ref{sec:testmodels}). For the instrument scenario \chandra we convolved the with a narrow Gausssian shaped PSF of width $\sigma=2$~pix. For the scenario \xmm we used a Gaussian shaped PSF with a width of $\sigma=2$~pix and a five times increased effective are.
        All images are of size $128{\times}128$.  The dashed red and magenta boxes depict $32{\times}32$ and $64{\times}64$ regions of interest.  Note that the boxes are displaced to the lower left corner in Case B, and leftwards in Case D.
        convolved with a diffuse PSF (Gaussian with $\sigma=6$~pix), \todo{add back boxes if we have time}.
        }
        \label{fig:simulated_models_chandra}
    \end{figure*}

    Constant background levels of $\lambda_{Bkg}= 0.001, 0.01 \textrm{and} 0.1 \textrm{counts/pixel}$.

    \begin{table*}[t]
    \label{tab:scenarios}
      \centering
      \begin{tabular}{l|cr}
        \hline
        Scenario & Description & Parameters \\
        \hline
        A1 & A grid of point sources & x, y \\
        B1-3 & Extended Gaussian shaped source surrounded by point sources & $\sigma$ \\
        C1-3 & Point Sources on top of a disk shaped source, jet like line features & $r_{\mathrm{Disk}}$ \\
        D1-5 & A spiral-arm structure surrounded by point sources and a large ring like feature & 9 \\

@@

@@

        \item \textbf{Jolideco (Patch, GLEAM v0.1)}: \jolideco with a patch prior learned on GLEAM data, described in Section~\ref{sssec:gleam-radio-data}. It also uses the standardised patch normalisation as described by Equation~\ref{eq:patch-norm}. We use cycle spinning, a stride of 4 pixels and 2000 \textit{training epochs}. Given 5 observations this results in 10.000 iterations. Which optimizes "to the end" until there is no relevant change in the posterior log-likelihood. In addition we use and upsampling factor of 2.

@@

    We ran the comparison on the whole parameter grid of methods, source scenarios, background levels and instrument scenarios. However in the following we only discuss a subset of the results, which we consider the most relevant. The complete set of results is available at \url{https://jolideco.github.io/jolideco-comparison}.

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \script{comparison-scenarios.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of different deconvolution methods and prior assumptions for different selected source scenarios, as described in Section~\ref{subsec:test-datasets}. The simulation of the data for this figure used a fixed background level of $\lambda_{Bkg} = \qty[mode = math]{0.01}{cts/pix}$, a uniform exposure and a Gaussian PSF of width $\sigma_{PSF} = \qty[mode = math]{2}{pix}$, corresponding to the \enquote{Chandra} instrument scenario. The leftmost column shows the simulated counts data and the second most column from the left the underlying ground truth to compare. The other columns show the reconstructed flux by the different methods and prior assumptions for \jolideco. To enhance weak structures the images use an $\arcsinh$ stretching with a scale parameter of $a=0.02$. The stretching is the same for all images. The methods are described in detail in Section~\ref{subsec:methods}. A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of methods and prior assumptions on the different source scenarios. Figure \ref{fig:comparison-scenarios} shows the results for the selected subset of scenarios A1, B3, C3 and D4.

@@

    For scenario B3 all method recover the point sources and the central extend source. However large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected the uniform prior decompose the emission into many point sources, neglecting the spatial correlation between them. \textbf{Pylira} shows slightly improved results with the gaps between the bright emission peeks being more filled, however still lacks a convincing result into reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. The GLEAM v0.1 patch prior reconstructs the sources almost perfectly, visually almost indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a visible extension in the surrounding point sources.

    Scenario C3 challenges the methods to reconstruct a weak extended source under four brighter point sources. While again all methods reconstruct the point sources well, the visual differences are large for the weak underlying extended emission. The uniform prior with 1000 iterations decomposes the source into a grid of almost equidistant point sources resembling only the outer boundary of the original source. \textit{Pylira} shows a similar tendency but already improves the reconstruction of the extended source by leaving more emission in between the bright emission spots. The best results are achieved by the patch priors, which reconstruct a plausible extended source with approximately uniform brightness.

    On the jet like features the patch prior does not show a preferred direction or bias in reconstruction and preserves the linear correlation between the pixels.

    \jolideco shows excellent performance of a variety of morphological shapes. 
    it converges to the reference solution for very high signal to noise. We find the prior learned from GLEAM data to be a very good compromise between teh ability to separate nearby point sources and the ability to reconstruct the spatial correlation of extended sources.

    \subsubsection{Comparing Signal-to-Noise Scenarios}
    %
    \begin{figure*}
        \script{comparison-signal-noise.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-signal-noise.pdf}
            \caption{
                Comparison of the \jolideco method with the GLEAM v0.1 prior for different background levels and signal strength represented by scenario D1-5. The images are grouped by background levels with each column group showing the counts data in the left column and the reconstruction in the right column. The rows show the scenarios D1-5 as described in Section~\ref{subsec:test-datasets}. The background levels correspond to \qty[mode = text]{0.001}{cts / pixel}, \qty[mode = text]{0.01}{cts / pixel} and \qty[mode = text]{0.1}{cts / pixel}. The complete results for the simulations and additional information are available at \url{https://jolideco.github.io/jolideco-comparison}. \todo{the background handling does not yet seem right...maybe we should let the background norm free}
            }
            \label{fig:comparison-signal-noise}
        \end{centering}
    \end{figure*}
    %
    As a second experiment we explored the dependency of the results on the background level and signal strength. Figure~\ref{fig:comparison-signal-noise} shows the results for the \jolideco method with the GLEAM prior in the scenario D1-5. The background level was varied between

    As expected the reconstruction quality improves with signal strength. This includes both the variation of the background as well as signal strength. \todo{sensitivity scales as signal / sqrt(Bkg)? } Visually the method converges to the ground truth for very high signal-to-noise ratio. Once the signal strength becomes very large there is no dependency of the result on the background level anymore (especially valid for scenario D5)

    We observe the method struggles to reconstruct a useful signal in the case of very high background (scenarios D1-3). 
    \todo{this is somewhat surprising as visually the signal can be seen in the data...at least for D2 and D3}.
    \todo{We should give a single number for the S/N, how to compute? Per source, like spiral ring and point sources...}
